---
title: "Accidental Gun Deaths and Injuries for Children and Teens"
author: "Teresa Do"
output: html_document
---

This project investigates accidental gun deaths and injuries for children and teens based off of data provided by the [Gun Violence Archive](https://www.gunviolencearchive.org/).

The goal of this project is to investigate the states and cities/counties that include the most incidents of accidental child and teen deaths by firearms, both in total counts and per capita.

# Set up

Loading libraries.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(tidycensus)

# disable scientific notation
options(scipen = 999)
```

# Load data from Gun Violence Archive

```{r}
#| label: load-data
#| message: false
#| warning: false

# for now, I'm focusing just on accidental child deaths
# I can also take a look at all child deaths
child_deaths <- read_csv("../data/2026-06-02-accidental-child-deaths.csv") |> clean_names()
teen_deaths <- read_csv("../data/2026-06-02-accidental-teen-deaths.csv") |> clean_names()

child_deaths
teen_deaths
```

We can also combine these into one data frame by binding the two tibbles together.

```{r}
#| label: bind-data
#| message: false
#| warning: false

child_deaths$teen_or_child <- "Child"
teen_deaths$teen_or_child <- "Teen"

all_deaths <- rbind(child_deaths, teen_deaths) |>
  unique()

# checking to see if any child incidents and teen incidents overlap
all_deaths |>
  get_dupes()

all_deaths
```

# Total counts of deaths

In this portion, I'll just take a look at the total counts of deaths for teens and children across states and cities/counties.

> NOTE: Need to look into what Gun Violence Archive defines the categories or "Teen" and "Child" because teen is a bit ambiguous

## Teens

The Gun Violence Archive has a [pre-queried page](https://www.gunviolencearchive.org/accidental-teen-deaths?page=62) for accidental deaths of teens.

Let's take a look at the total sum of accidental teen deaths across states.

```{r}
#| label: teen-deaths-state
#| message: false
#| warning: false

teen_by_state <-teen_deaths |>
  group_by(state) |>
  summarize(
    deaths = sum(victims_killed)
  ) |>
  arrange(deaths |> desc())

teen_by_state
```

Based off of just total counts, Texas has the most accidental teen deaths by firearm. We will also take a look how this stacks up per capita, taking into account that Texas is a big state.

I'm also interested in looking at this on the city/county level.

```{r}
#| label: teen-deaths-county
#| message: false
#| warning: false

teen_deaths |>
  group_by(state, city_or_county) |>
  summarize(
    deaths = sum(victims_killed)
  ) |>
  arrange(deaths |> desc())
```

Again, Texas is in first place with Houston, which is the fourth-most populous city in the country.

## Children

Do the same for accidental children deaths by state.

```{r}
#| label: child-deaths-state
#| message: false
#| warning: false

child_by_state <- child_deaths |>
  group_by(state) |>
  summarize(
    deaths = sum(victims_killed)
  ) |>
  arrange(deaths |> desc())

child_by_state
```

When it comes to total counts of accidental children deaths by firearms, Tennessee comes up at 27 deaths. Again, I will also take a look at how this stacks up per capita.

Looking now at city/county level.

```{r}
#| label: child-deaths-county
#| message: false
#| warning: false

child_deaths |>
  group_by(state, city_or_county) |>
  summarize(
    deaths = sum(victims_killed)
  ) |>
  arrange(deaths |> desc())
```

Memphis, Tennessee has 12 accidental child deaths by firearms but is also a pretty big city.

# Per capita deaths

We want to be able to examine these values in the context of the total population of an area. First, let's figure out what [variables](https://walker-data.com/census-r/an-introduction-to-tidycensus.html#searching-for-variables-in-tidycensus) to use with the `tidycensus` package.

```{r}
#| label: tidycensus-variables
#| message: false
#| warning: false

# ask about using decennial vs. acs vs. get_estimates()
vars <- load_variables(2024, dataset = "acs5", cache = TRUE)
# View(vars)
```

`B01001_001` is the code for population estimates for a block group.

Using this variable, we can find the population by state and then bind these population values to the data frames containing deaths by state for both teens and children.

```{r}
#| label: state-pop
#| message: false
#| warning: false

population <- get_acs(
  geography = "state",
  variables = "B01001_001",
  year = 2024,
  survey = "acs5"
) |>
  clean_names() |>
  select(
    geoid,
    state = name,
    pop_estimate = estimate
  )

population
```

I can join these using state name to be able to calculate for teens and children. 

## Teens

Let's calculate the accidental teen gun deaths per capita by state. Combine the two data frames with a left_join (because Puerto Rico is not included in Gun Violence Archive)

```{r}
#| label: combine-teen-w-pop
#| message: false
#| warning: false

teen_by_state_w_pop <- teen_by_state |>
  left_join(
    population,
    by = join_by(state)
  )

teen_by_state_w_pop
```

Now, calculate per 100,000 rates for accidental teen deaths by guns per state. We are choosing to do per 100,000 because the rates are so small.

```{r}
#| label: teen-deaths-per-100000
#| message: false
#| warning: false

teen_by_state_per_100000 <- teen_by_state_w_pop |>
  mutate(
    per_100000 = deaths / pop_estimate * 100000
  ) |>
  arrange(per_100000 |> desc())
```

Based on per capita calculations, Mississippi, Louisiana and Alabama actually have the highest rates of accidental teen deaths by guns.

Export for visualization in Datawrapper.

```{r}
#| label: export-teen-data
#| message: false
#| warning: false

teen_export <- teen_by_state_per_100000 |>
  select(
    state,
    per_100000
  )

top_20_teen <- teen_export |> head(20)

teen_export |>
  write_csv("../data/2026-06-03-teen-per-100000.csv")

top_20_teen |>
  write_csv("../data/2026-06-03-teen-top-20.csv")
```

## Children

Now, let's calculate the accidental child deaths by gun per 1,000,000 people. First, combine the population data frame with our existing child deaths data frame.

```{r}
#| label: combine-child-w-pop
#| message: false
#| warning: false

child_by_state_w_pop <- child_by_state |>
  left_join(
    population,
    join_by(state)
  )

child_by_state_w_pop
```

Now, we can calculate the per 1,000,000 rate. We chose to calculate this as per 1,000,000 people since the number of child deaths are so much lower than number of teen deaths.

```{r}
#| label: child-deaths-per-100000
#| message: false
#| warning: false

child_by_state_per_1mil <- child_by_state_w_pop |>
  mutate(
    per_1000000 = deaths / pop_estimate * 1000000
  ) |>
  arrange(per_1000000 |> desc())

child_by_state_per_1mil
```

Tennessee still comes in first with around 4 accidental child gun deaths per 1,000,000 people.

Export for visualization in Datawrapper.

```{r}
#| label: export-child-data
#| message: false
#| warning: false

child_export <- child_by_state_per_1mil |>
  select(
    state,
    per_1000000
  ) |>
  head(20)

child_export |>
  write_csv("../data/2026-06-03-child-per-1mil.csv")
```

