---
title: "Accidental Deaths of Children and Teens Involving a Firearm - Teresa Do"
author: "Wells"
date: "2026-05-25"
output: html_document
---

The Trace - Gun Violence Analysis
For Teresa Do
Teresa will compile and analyze accidental gun deaths and injuries for children and teens.


Gun Violence Archive data
(The starter code has older data. You should update the downloads)
https://www.gunviolencearchive.org/children-killed
https://www.gunviolencearchive.org/accidental-teen-deaths
The Gun Violence Archive is an online archive of gun violence incidents collected from over 7,500  law enforcement, media, government and commercial sources daily in an effort to provide near-real time data about the results of gun violence. GVA is an independent data collection and research group with no affiliation with any advocacy organization.



# load libraries
```{r}
install.packages("tidycensus")
library(tidyverse)
library(tidycensus)
library(janitor)
```

#load data from Gun Violence Archive
```{r}
#Looks like you should update the downloads

#https://www.gunviolencearchive.org/children-killed
children <- read_csv("../data/AccidentalChildDeaths.csv") |> 
     clean_names() 
#https://www.gunviolencearchive.org/accidental-teen-deaths
teens <- read_csv("../data/AccidentalTeenDeaths.csv") |> 
     clean_names()

```

#Combine into one dataframe
```{r}
children$teen_or_child <- "child"
teens$teen_or_child <- "teen"
all <- rbind(children, teens)

```

## Deaths By State for Children

```{r}
options(scipen = 999)  # Disable scientific notation globally

children |> 
group_by(state) |> 
  summarize(
    Deaths = sum(victims_killed)
  ) |> 
  arrange(state)

```

## Deaths By State for Teens
```{r}

teens |> 
group_by(state) |> 
  summarize(
    Deaths = sum(victims_killed)
  ) |> 
  arrange(state)
```

## Arrests for One City or County for Deaths Involving Children
```{r}
children |> 
  filter(city_or_county == "Baton Rouge") |> 
  group_by(city_or_county) |>
  summarize(
    Arrests = sum(suspects_arrested)
  ) |> 
  arrange(desc(Arrests))

```

## Deaths for One State and City or County for Deaths Involving Teens
```{r}
teens |> 
  filter(state == "Maryland") |> 
  filter(city_or_county == "Baltimore") |> 
  group_by(city_or_county) |>
  summarize(
    Deaths = sum(victims_killed)
  ) |> 
  arrange(desc(Deaths))

```


The CDC Wonder Database
https://wonder.cdc.gov/controller/datarequest/D158;jsessionid=B8BD647539274A8EF41B01B03A5C


The Underlying Cause of Death database contains mortality and population counts for all U.S. counties. Data are based on death certificates for U.S. residents. Each death certificate identifies a single underlying cause of death and demographic data. The number of deaths, crude death rates or age-adjusted death rates, and 95% confidence intervals and standard errors for death rates can be obtained by place of residence (total U.S., region, state and county), age group (single-year-of age, 5-year age groups, 10-year age groups and infant age groups), race, Hispanic ethnicity, sex, year, cause-of-death (4-digit ICD-10 code or group of codes), injury intent and injury mechanism, drug/alcohol induced causes and urbanization categories. Data are also available for place of death, month and week day of death, and whether an autopsy was performed.
